---
title: "Data Analysis"
output: html_notebook
---

First, I'll set up my script
```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
```

Now, I will import my two (cleaned up) datasets 
```{r}
regen <- read.csv(here("Data", "Processed", "correct_regen.csv"), stringsAsFactors = TRUE)
density <- read.csv(here("Data", "Processed", "correct_density.csv"), stringsAsFactors = TRUE)
```


## Statistical Test 1: Canopy Density and Regeneration Count (linear regression)

Before I begin the data analysis workflow, I need to manipulate my data because these two variables are found in different dataframes.

A. First I will summarize the regen data to get the a dataframe that includes Forest, Plot, and the number of total regen (not distinguished by species) for that plot
```{r}
regen_by_plot <- regen %>% 
  group_by(Forest, Plot_num) %>%
  summarize(num_seedlings = mean(Regen_count))
```

B. Next I will summarize the density data to get a dataframe with the average canopy density (densiometer method) for each plot
```{r}
avg_density <- density %>%
  group_by(Forest, Plot_num) %>%
  summarize(density = mean(Densiometer))
```


C. Now I will create a new column in each of the dataframes I created above that combines the forest and plot so that each site has a unique ID I can use for combining the dataframes
```{r}
regen_by_plot$ID <- paste(regen_by_plot$Forest,regen_by_plot$Plot_num)

avg_density$ID <- paste(avg_density$Forest,avg_density$Plot_num) #that worked so well! 
```

D. Now I will join the dataframes
```{r}
combined_data <- regen_by_plot %>% full_join(avg_density, by = c("ID")) %>%
  select(-c(Plot_num.x, Forest.x)) %>%
  rename(c("Forest"="Forest.y", "Plot_num" = "Plot_num.y")) #looks like everything I need is in this dataframe, and there are no duplicate collumns 
```

I will move into the analysis workflow 

### 1. I will plot a histogram of each variable I am comparing

Canopy density first. 
```{r}
ggplot(combined_data, aes(x = density)) + 
  geom_histogram() +
  theme_bw() #I do not see one lone outlier, this looks good
```

Now I will look at the Regen_count variable 
```{r}
ggplot(combined_data, aes(x = num_seedlings)) +
  geom_histogram() +
  theme_bw() #In my data exploration I looked at this variable and determined the data were legit, so even though it looks like there is an outlier I feel good to move ahead.
```

### 2. I will plot the variables together
```{r}
ggplot(combined_data, aes(x = density, y = num_seedlings)) +
  geom_point() +
  theme_bw()
```

### 3. Guess relationship: Based on the data, it appears there could be a weak positive relationship between canopy density and the number of seedlings.

Total rise: 14
Total run: 58 to 94, so 44
Slope: 15/44 = 0.34  
Y intercept: 0ish


What statistical test to use: I will use a linear regression since the predictor variable (canopy density) is continuous, and the response variable (regeneration) is also continuous. 

### 4. Create the Model
```{r}
regen_dens_mod <- lm(num_seedlings ~ density, combined_data)
```


### 5. Check Assumptions
```{r}
autoplot(regen_dens_mod, smooth.colour = NA)
```
The normal Q-Q graph looks good, nand indicates the data are fairly normally distributed, with some smaller and larger values deviating. The Residuals vs. fitted graph indicates relatively equal variance. I am comfortable moving forward.  

### 6. Interpret Model:

```{r}
anova(regen_dens_mod) #asking for standard output of an anova, not calculating an anova  
```
The P-value is 0.8527 which is large - accept null hypothesis (that there is no meaningful statistical relationship b/w regen and canopy density) 

I will also run the summary 
```{r}
summary(regen_dens_mod) 
```
Estimates R-squared to be 0.002375, which means only a 0.2% of variation in regen is explained by canopy density. 

I will not make a publication plot since no statistically significant findings were found. 



## Statistical Test 2: Canopy Density and Percent Bare Ground (linear regression)

### 1. First I will plot each variable

Canopy density first
```{r}
ggplot(density, aes(x = Canopyapp)) +
  geom_histogram() +
  theme_bw() #does not appear to be one lone outlier
```

Percent bare ground second
```{r}
ggplot(density, aes(x = Percent_bare)) +
  geom_histogram() +
  theme_bw() #no outliers
```

### 2. Now I will plot the variables together
```{r}
ggplot(density, aes(x = Canopyapp, y = Percent_bare)) +
  geom_point() +
  theme_bw()
```

### 3. Guess the relationship: From looking at the data, it appears there could be a weak positive relationship between canopy density and percent bare ground

Total rise: 100
Total run: 38 to 98 = 60
Slope: 100/60 = 1.67
Y intercept: 0ish?

Null Hypothesis: there is no significant statistical relationship between canopy density and percent bare ground

### 4. Create a model 

Because both the predictor (canopy density) and response (percent bare ground) are continuous, I will use a linear regression. 

I'll construct the model
```{r}
density_bare_ground_mod <- lm(Percent_bare ~ Canopyapp, density)
```

### 5. Check the assumptions 
```{r}
autoplot(density_bare_ground_mod)
```
Normal Q-Q plot: shows that most of the data points fit a normal distribution, with some of the higher and lower points deviating, but nothing excessive
Residuals vs fitted plot: shows that variance of residuals is fairly equal (slight waiver in line but nothing too drastic)
Based on the autoplot() results, I feel moderately comfortable moving ahead and assuming these data meet the linear model assumptions

### 6. Interpret the model
```{r}
anova(density_bare_ground_mod)
```
The P-value is 0.01603, which is fairly small. I reject the null hypothesis, and accept that canopy density has a significant effect on ground cover. 

I will also run a summary of the test
```{r}
summary(density_bare_ground_mod)
```

Slope is estimated to be -5 and slope as 0.7, which is fairly similar to what I predicted. 

### 7. Publication Plot
```{r}
ggplot(density, aes(x = Canopyapp, y = Percent_bare)) +
  geom_point(size = 2) +
  geom_smooth() +
  xlab("Canopy Density") +
  ylab("Percent Bare Ground")
  theme_bw() #adding the smoother provides a better visual for the slight positive relationship between canopy density and bare ground
```



## Statistical Test 3: Percent Ferns Between Kip and Donnerville (Generlized Linear Model)

Before I start the data analysis workflow, I need to filter my data to only get fern data from Kip and Donnerville. The reason I choose to look at these two forests is because during my exploratory data phase I saw there could be a difference between the amount of ferns in these two forests. 
```{r}
ferns <- density %>% 
  filter(Forest == "Kip" | Forest == "Donnerville")
```

 
### 1. Plot each variable

First, I'll plot the percent ferns
```{r}
ggplot(ferns, aes(x = Percent_ferns)) +
  geom_histogram() +
  theme_bw() #does not appear to be any lone outliers
```

Next, I'll double check the levels of the Forest variable to make sure there are no mistakes
```{r}
unique(ferns$Forest) #only 2 forests, good
```

### 2. Plot variables together
```{r}
ggplot(ferns, aes(x = Forest, y = Percent_ferns)) +
  geom_boxplot() +
  theme_bw() #looks like there is variation in the amount of ferns between forests
```
### 3. Guess relationship
From looking at the data, I predict there to be a significant difference between the amount of ferns at each forest. It appears there are more ferns in the Kip forest. 

Null hypothesis: there is no significant relationship between the forest and the amount of ferns present

### 4. Create model

I will fit a generalized linear model because I am dealing with a categorical predictor variable with more than two groups (Forest) and a continuous, yet bounded, (0-1) response variable (percent ferns)
```{r}
ferns_forest_glm <- glm(Percent_ferns ~ Forest, data = ferns, family = poisson())

#poisson family has a ___ link and assumes a poisson distribution
```

### 5. Check assumptions
```{r}
autoplot(ferns_forest_glm)
```
Normal Q-Q plot: it appears the data fit to the normal distribution with some deviation for small and large values. 

```{r}
summary(ferns_forest_glm)
```

Residuals vs fitted plot: shows that variance of residuals is fairly equal. 

Based on the autoplot() results, I will feel fairly ok with proceeding.   

### 6. Interpret the model 
```{r}
anova(ferns_forest_glm, test = "Chisq") #very small p-value!!
```

Also run the summary table
```{r}
summary(ferns_forest_glm) 
```

The small p-value leads me to reject my null hypothesis. There is a significant difference in the amounts of ferns between Donnerville and Kip. 

### 7. Now I will make a publication plot

```{r}
ggplot(ferns, aes(x = Forest, y = Percent_ferns)) +
  geom_boxplot() +
  geom_point(position = "jitter", size = 2, alpha = 0.5, color = "blue")+ #generate "noise" to show possible data points 
  theme_bw() +
  ylab("Percent Ferns") +
  xlab("Forest")
```


## Statistical Test 4: Regen count between Peavine and Degrasse (chi squared -- or GLM)

First I will filter my data to get regen count for just South Hammond and Degrasse forests. The reason I am choosing these forests is because I think it would be interesting to compare the amount of regen between a mostly coniferous forest (Degrasse) and a mostly deciduous forest (South Hammond). 

```{r}
levels(regen$Forest) #first double check levels

regen_chi_squared <- regen %>% 
  filter(Forest == "Degrasse" | Forest == "South Hammond") 


regen_chi_squared$Forest <- factor(regen_chi_squared$Forest) #So R relevels the Forest variable
levels(regen_chi_squared$Forest) #good
```


### 1. Plot the variables

Now I will plot a histrogram to look for any outliers
```{r}
ggplot(regen_chi_squared, aes(x = Regen_count)) +
  geom_histogram()
  #nothing looks to crazy, and these numbers seem correct considering what they represent  
```

Now I will manipulate the data before I plot again 
```{r}
total_regen <- regen_chi_squared %>% group_by(Forest) %>% summarise(total_regen = sum(Regen_count))
```

### 2. Plot varaibles together
```{r}
ggplot(total_regen, aes(x = Forest, y = total_regen)) +
  geom_bar(stat = 'identity', position = 'dodge') + #plotting just the means (identity)
  theme_bw() 
```

### 3. Guess relationship
From looking at the data, it appears there is a significant difference between the amount of regeneration found between the degrasse and south hammond forests. It seems there is more regen in south hammond. 

### 4. Create model
Now, because I am dealing with a categorical variable with two levels (Forest) and count data (regen) I will run a chi squared test. first I will fit the data to a matrix.
```{r}
regen.mat <- xtabs(total_regen ~ Forest,
                  data = total_regen)

regen.mat #shows a table
```


### 5. Check assumptions
Count data are assumed to have a normal distribution, so I am not going to check any assumptions here. 

### 6. Interpret the model

Now I will run the chi squared test
```{r}
regen.chi <- chisq.test(regen.mat) 

regen.chi #smallish p-value
```
Because the p-value is relatively small (less than 0.05), I will reject the null hypothesis. It seems like there is a slightly significant difference between the amount of regen in South Hammond and Degrasse, meaning the forest has an impact on the amount of regen present. 

### 7. Publication plot
```{r}
ggplot(regen_chi_squared, aes(x = Forest, y = Regen_count)) +
  geom_boxplot() +
  geom_point() +
  theme_bw() +
  ylab("Regen Count")
```










