---
title: "Data Exploration"
output: html_notebook
---

## Goals for this script:

This is where I will explore my data and check for errors in data entry. I will take an overview look at the data, and then plot individual variables. 


First, I'll set up my workspace
```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
```

Now I'll pull in my data
```{r}
regen <- read.csv(here("Data", "Regen_data_summer22.csv"))
density <- read.csv(here("Data", "Ground_cover_canopy_density_summer22.csv"))
```


Since I have two dataframes, I will perform data anlaysis on them one at a time. I'll start with the regen dataset first

## Regen Data Exploration

A. First I want to take a look at the structure of the data
```{r}
str(regen) #I see that some variables have been imported as characters, I want to change them to factors
```

I will reimport the data and assign characters to factors upon import
```{r}
regen <- read.csv(here("Data", "Regen_data_summer22.csv"), stringsAsFactors = TRUE)

str(regen) #ahh, much better
```

B. Now I want to look at a more in depth summary of the data
```{r}
summary(regen) #I am not seeing anything suspicious as far as outliers, but plotting help will confirm that 
```

C. Now I want to plot histograms of each continuous variable that I might be working with and visualize the mean and range for each 

1. Percent clouds
```{r}
ggplot(regen, aes(x = Percent_clouds)) +
  geom_histogram() #don't see anything suspicious
```
From looking at this histogram, it appears the mean percent clouds is about 50, with a range from 0-100. 


2. Regen count
```{r}
ggplot(regen, aes(x = Regen_count)) +
  geom_histogram() #Hmm, there could be an outlier. Looking at the data I see that the highest value occurs in Kip plot 1 with 24 red maple seedlings. Because I know the plot and that there are quite a few red maples, I don't think this is misentered data. 
```

From looking at this histogram, it appears the mean regen count is around ~5, and the range is about 0-24. 

You may also want boxplots of these variables by forest for example.


D. Now I want to look at the levels of some of the factor variables I will use to make sure there is no misentered data

1. Forest
```{r}
levels(regen$Forest) #oops, looks like Donnerville has a spelling issue. I will fix it. I want there to be only 6 levels and right now there are 7

bad <- which(regen$Forest == "Donerville")
regen$Forest[bad] <- "Donnerville"
unique(regen$Forest) #that looks better!
```

2. Regen species
```{r}
levels(regen$Regen_species) #don't see any mistakes
```

3. Regen type
```{r}
levels(regen$Regen_type) #only two types, good
```

E. Now I want to plot boxplots with some factor variables 

1. Forest and regen_count
```{r}
ggplot(regen, aes(x = Forest, y = Regen_count)) +
  geom_boxplot() #again, I see that high value that occurs on the kip trail. However, since I remember collecting the data and that there were a lot of red maples there I am not concerned
```

2. Regen species and regen count
```{r}
ggplot(regen, aes(x = Regen_species, y = Regen_count)) +
  geom_boxplot() +
  coord_flip()
#looks like there are lots of juneberry and musclewood present, and quite a few red maple (with 24 at that one site)
```

3. Regen type and regen count
```{r}
ggplot(regen, aes(x = Regen_type, y = Regen_count)) +
  geom_boxplot() +
  coord_flip() #can see that the understory in these plots is almost entirely seedlings, with only a few saplings
```

Lastly, I will save a new version of my regen data to save the changes I made (e.g. fixing spelling errors in the Forest variable) in the Processed data folder
```{r}
write.csv(regen, here("Data", "Processed", "correct_regen.csv"))
```

Nice data exploration!

## Density and Ground Cover Data Exploration 

A. First I'll look at the structure of the data
```{r}
str(density) #got some character variables that I want to change to factors
```

I will re-import with stringsAsFactors()
```{r}
density <- read.csv(here("Data", "Ground_cover_canopy_density_summer22.csv"), stringsAsFactors = T)

str(density) #good
```

B. And now for a summary of the dataframe
```{r}
summary(density) #nothing seems too out of order 
```


C. Now I want to plot histograms of each continuous variable that I might be working with and visualize the mean and range for each 

1. First I'll plot percent bare ground (these next few variables are data taken from ground cover plots)
```{r}
ggplot(density, aes(x = Percent_bare)) +
  geom_histogram() #doesn't appear to be any outliers
```

From looking at the plot, I would guess the mean for percent bare ground is about 40, and the range is from around 0-95. 

2. Percent non-vascular
```{r}
ggplot(density, aes(Percent_nonvascular)) +
  geom_histogram() #there is no one lone outlier, and nothing greater than 100 (these are percents!) so seems reasonable
```

From looking at the graph I predict the mean of percent non vascular ground cover to be around 10, with a range from around 0-75. 

3. Percent ferns
```{r}
ggplot(density, aes(x = Percent_ferns)) +
  geom_histogram() #doesn't appear to be one lone outlier and nothing beyond 100
```

I predict the mean percent ferns to be around 10, with a range from about 0-35. 

4. Percent angiosperms
```{r}
ggplot(density, aes(x = Percent_angiosperms)) +
  geom_histogram() #doesn't appear to be any outliers
```

From looking at the plot I guess the mean percent angiosperms in the ground-cover plots to be about 35, with a range from about 2-75. 

5. Regen densiometer (a measure of canopy density, so a percent, taken above a regen plot)
```{r}
ggplot(density, aes(x = Regen_densiometer)) +
  geom_histogram() #doesn't appear to be any outliers
```

I predict the mean to be about 87, with a range from about 58-97. 

6. Regen canopyapp (another canopy density measurement above a regen plot)
```{r}
ggplot(density, aes(x = Regen_canopyapp)) +
  geom_histogram() #doesn't seem to be any outliers
```

I predict the mean to be about 85, with a range from about 53-96. 

7. Densiometer (randomized canopy density reading from the forest plot)
```{r}
ggplot(density, aes(x = Densiometer)) +
  geom_histogram() #doesn't appear to be any outliers
```

From looking at the graph, I predict the mean to be about 87, with a range from 53-100. 

8. CanoppyApp (rangomized measurement of canopy density within the forest plot)
```{r}
ggplot(density, aes(x = Canopyapp)) +
  geom_histogram() #doesn't appear to be any outliers
```

I predict the mean to be about 82, with a range from 40-98. 

D. Now I want to look at the levels of some of the factor variables I will use to make sure there is no misenterd data
```{r}
levels(density$Forest) #looks good
```

E. Now I want to plot a couple of boxplots that show some of the relationships between variables (considering a few variables I plan to use in my analysis)

1. Canopy density (I will just do the CanopyApp method for this graph) and Forest
```{r}
ggplot(density, aes(x = Forest, y = Canopyapp)) +
  geom_boxplot() #looks like there could be some fun things happening...e.g. maybe lower canopy density in Degrasse than Donnerville? (possible statistical test!)
```

2. Percent ferns and forest
```{r}
ggplot(density, aes(x = Forest, y = Percent_ferns)) +
  geom_boxplot() #looks like there could be differences here as well
```

3. Forest and percent bare ground
```{r}
ggplot(density, aes(x = Forest, y = Percent_bare)) +
  geom_boxplot() #nice!
```

And again, I will save this dataframe to the Processed data folder to use later in my analysis
```{r}
write.csv(density, here("Data", "Processed", "correct_density.csv"))
```



From working through this data exploration I feel confident in how my data are entered, and the histograms of variables seem to make sense. Additionally, from examining a few preliminary box plots, I can see potentially interesting results from statistical tests I plan to conduct in the next section of the Capstone Data Analysis Project. 

Nice work!


